Regression Cube Analysis of Cohort Study Data

### 2015-03-29 draft.tex
revised using Sylvias notes; shortened related work

%\com{New Solution}.
%%
We propose an analysis approach of epidemiological data sets by incorporating all features in an exhaustive regression-based analysis.
%%
This approach combines all \emph{independent features} with respect to a \emph{target feature} and provides a visualization, which reveals insights into the data by highlighting relationships.
%%
A 3D-visualization of all combinations of two to three independent features towards a target acts as an overview of the whole data set, the \emph{Regression Cube}.
%%
Slicing through the \emph{Regression Cube} allows for the detailed analysis of features towards the target disease.
%%
Expert knowledge about disease-specific hypotheses can be included into the analysis by adjusting the regression model formulas.
%%
Furthermore, the influences of features can be assessed using a difference view comparing different calculation results.
%%
%\com{Validation \& Results}.
%%
We applied our \emph{Regression Cube} method to a hepatic steatosis data set to reproduce results from a data-mining driven analysis.
%%
A qualitative analysis with three domain experts was conducted on a breast fat data set.
%%
We were able to derive new hypotheses about relations between breast fat density and breast lesions towards breast cancer.
%%
%\com{Implications}.
%%
%With our work, we present a visual overview of epidemiological data with the \emph{Regression Cube}, which allows for the first time an interactive regression-based analysis of large feature sets with respect to a disease.
} % end of abstract

@@

\maketitle
%%
Epidemiology aims to characterize health and disease conditions in defined populations (\emph{cohorts}).
%%
Insights about risk factors allow to characterize disease-specific high-risk groups and act as important diagnostic key figures \cite{Fletcher2012}.
%%
Furthermore, they can be used to derive recommendations regarding a healthy lifestyle or to provide information about wide-spread diseases.
%%
During the standard workflow, physicians transform observations into hypotheses, which are depicted using epidemiological features and then assessed using regression analyses.

@@

Testing features for association with diseases using regression analysis is one of the most important epidemiological tools.
%%
Using it to assess the statistical resilience of a hypothesis rarely involves \emph{more than three features} due to the required subject count.
%%
% missing overview techniques ist zu hart; es gibt ja techniken (eine scatterplot matrix zb.)
Due to the amount of data and only limited overview techniques, possible correlations may be missed.
%%
Explorative analyses and overview visualizations of the data set as presented by Klemm et al. \cite{Klemm2014VIS} are not tailored to a specific target variable and mostly highlight correlations between variables, which are known to the domain expert (e.g., correlation between body size and spine shape).
%%
We incorporate the regression analysis, which is familiar to the domain experts, into overview visualizations to support a hypothesis-free analysis or an analysis towards a specific disease or hypothesis.
%%
This is achieved by providing template regression formulas, which are applied to all potential variable combinations.
%%
Since the notation is familiar to epidemiologists, they can rapidly include their domain knowledge into the analysis process.
%%
Difference views between regression formulas allow to assess the influences of individual variables on the process.

Our contributions are:
\begin{itemize}
	\item An overview visualization technique representing feature interactions using target features.
	\item Visualization techniques, which incorporate overview visualizations of all regression analyses at once as well as details on demand techniques for in-depth investigations of feature relationships.
	\item Freely adjustable regression formulas provide a simple, yet powerful way to adjust the regression analysis to specific hypotheses about the data.
	\item The analysis of confounding variables by providing comparison views between different formula results.
%	\item The open and web-based approach of the system allows for analysis of any data using the presented method.
	\item The expansion and adaption of our method to any data with our open and web-based system.
\end{itemize}
%%
The term Regression Cube has been used for the detailed analysis of one or a small number of regression models \cite{Chan, Ahmadi}.
%%
They extend 2D scatter plots to display detailed information about a regression model.
%%
In behalf of that, we declare our visualization technique for comparing and analyzing a large number of different regression models with different input features as Regression Cube.
%%
\section{Epidemiological Background} \label{sec:Background}
%%
This section covers the epidemiological workflow and requirements.
%%
\subsection{Epidemiological Workflow} \label{EpidemiologicalWorkflow}
Epidemiological research is performed by experts from different academic disciplines, such as physicians, statisticians and medical computer scientists focusing on biometrics and image segmentation.
%%
Their goal is to derive disease-specific risk factors by assessing epidemiological features using statistical methods.
%%
As described by Thew et al. \cite{Thew2009} the epidemiological workflow is divided into these different steps:
\begin{itemize}
	\item Clinicians make observations in the daily practice, which are translated into hypotheses.
	\item Epidemiologists compile a list of variables depicting the hypothesis and include confounding variables.
	\item Statisticians assess the association of the derived features towards the investigated disease.
\end{itemize}
%%
Relative risks can be determined if a statistical resilient association of features towards a condition is extracted.
%%
They indicate the per-subject chance of developing the disease.
%%
Reproducibility of results is an epidemiological key requirement and guides all analyses steps.
%%
Statistical programs, such as \texttt{SPSS}, are used to analyze the data using the classical sequential epidemiological workflow. Hence, images are mostly used for the communication of results, rather than for providing insight into the data.
%%
An alternative data-driven approach is described as follows.

\paragraph{Data Driven Hypothesis-Generation-based Analysis.}
We already described an Interactive Visual Analysis approach for image-centric cohort study data, which connects to the variable listing step \cite{Klemm2014VIS}.
%%
Their methods aim to derive hypotheses through analysis of the data and observations about previously unknown feature correlations.
%%
Employing \emph{hypotheses generation} requires overview visualizations of feature correlations, which are not supported by standard statistical processors.

\subsection{Epidemiological Data} \label{sec:EpidemiologicalData}
%%
%\com{Bernhard: Due to overlaps to VIS'14 paper in this section, this can be shortened. E.g. "More discussion of data types and characteristics can be found in [...]"}
%%
\emph{Cohort studies} are a major epidemiological tool for gathering epidemiological features.
%%
They yield a highly heterogenous and incomplete information space.
%%
Subject data are collected either towards a specific disease or with the widest range possible.
%%
The latter allows the data set to be assessed towards different diseases.
%%
The feature space comprises information about lifestyle, somatometric variables, medical parameters, genetic data as well as medical image data.
%%
These features are derived through different modalities, such as questionnaires, medical examinations or laboratory analyses.
%%
Many features are sparse, such as follow up questions about a medication or treatment of a certain disease.
%%
Other features are exclusive for a sub-group, such as women-specific questions, e.g., number of born children or period status.
%%
%\com{Restriction of amount of data because it must not be triangulated and because the ethics committees are very restrictive}
\\\\
%\paragraph{Data Types.}
%%
Medical status variables or lifestyle factors are often of \emph{dichotomous} (binary) type.
%%
The data space comprises continuous variables (somatometric variables, such as body weight) as well as categorical variables (e.g. graduation type).
%%
The data heterogeneity has to be taken into account when the analysis method is chosen.
%%
Continuous data are often discretized (e.g. 10 year steps for age) to equalize the variable types and to simplify the method selection.
%%
However, this should be avoided, since it reduces the information space and also introduces a new information bias, as assumptions are modeled through the discretization.
%%

Medical image data is also acquired and analyzed in modern cohort studies.
%%
We incorporate image-derived features, but do not put focus on analyzing image data.
%%
More discussion of data types and characteristics can be found in \cite{Klemm2014VIS, Toennies2015, Preim2014}.

%\paragraph{Image Data.}
% \com{Can probably be cut (reference to Klaus' Paper for more details on this) (see Bernhards comment above)}
% %%
% Since the Rotterdam study, many modern cohort studies include medical image data.
% %%
% For ethical reasons, the imaging modalities usually not include ionizing radiation.
% %%
% The image quality is often inferior to clinical standards, which is a tradeoff between time and cost \cite{Preim2014}.
% %%
% These data are hard to analyze as they require segmentations to outline structures of interest.
% %%
% This process is prone to inter- and intra-observer variability when carried out manually.
% %%
% Automatic or semi-automatic solutions are a remedy for this problem, but are costly and need to be tailored to the structure of interest \cite{Toennies2015}.
% %%
% Image-derived variables are either dichotomous variables indicating a medical finding by a radiologist or continuous, describing the shape of segmented structures (e.g. volume or diameter).

\paragraph{Confounding Features.}
Features influencing the exposure as well as the outcome of an analysis are called confounders.
%%
The analysis model has to be adjusted by normalizing all included features towards the confounding feature.
%%
\emph{Age} is included as confounder in almost any epidemiological analysis, since most diseases (such as different cancer types) are more likely with increasing age.
%%
It also influences the general body condition and thereby almost all features acquired through cohort studies.
%%
Another important confounder is \emph{Gender}.
%%
Other confounders have to be selected by epidemiologists specific to the investigated condition.

@@

\paragraph{Regression Analysis Notation.} Regression formulas are usually denoted as follows:
%%
\begin{equation}
Dependent \sim Independent_1 + ... + Independent_n
%Dependent \sim \sum_{i=1}^n Independent_i
\label{eq:RegressionNotation}
\end{equation}
%%
The most used regression operators comprise:
%%
\begin{itemize}
	\item \fbox{$+,-$} inclusion/exclusion of the variable,
	\item $:$ inclusion of interactions between the variables (e.g. $x:y$),
	\item $*$ inclusion the variables as well as their interactions (e.g. $x*y$)
	\item $|$ (conditioning) inclusion of variable x, given y (e.g. $x|y$)
\end{itemize}
%%
The type of the target feature restricts the regression type.

\paragraph{Linear Regression for Continuous Target.} The basic type is the linear regression, creating a linear map from the space comprising the \emph{independent} features to the \emph{dependent} features.
%%
The \emph{dependent} variable has to be of a continuous type.
%%

\paragraph{Logistic Regression for Dichotomous Target.} Logistic regression implies a dichotomous target variable.
%%
The target is described by fitting a logistic function.
%%
Logistic models as opposed to linear models do not allow for extracting an $R^2$ goodness of fit value.
%%
Therefore, pseudo-$R^2$ values are extracted, such as the \emph{Nagelkerke $R^2$}, which mimics the behavior of the $R^2$.
%%
\emph{Nagelkerke $R^2$} cannot be compared to $R^2$ values extracted from linear regression model.

@@

\subsection{The Study of Health in Pomerania (SHIP)}
The \texttt{SHIP}, located in Northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%%
It does not focus on a specific disease, making the data set open for many diseases.
%%
Unique for the \texttt{SHIP} is the acquisition of medical image data per cohort subject.
%%
A second cohort, \texttt{SHIP-TREND} was started in 2012.
%%
Data for both cohorts are examined in a 5-year time span.
%%
New parameters are added in each iteration, extending the range of investigated diseases.
%%
For the latest two acquisitions (\texttt{SHIP-2} and \texttt{SHIP-TREND-0}), MRI scans are included into the cohort \cite{Hegenscheid2009, Ivanovska2014}.
%%
However, most examinations occur in all stages and are performed according to the same instructions to enable comparisons over the different stages of the study ("\emph{method loyality}").
%%
%Starting in 1997 with a cohort of 4,308 subjects, the \texttt{SHIP}, located in Northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%After the pioneering Rotterdam study (started in 1990), several MR imaging study initiatives were initiated.
%
%They slightly differ in clinical focus, acquired data and epidemiological research questions.
%
%Starting in 1997 with a cohort of 4,308 subjects, the \texttt{SHIP}, located in Northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%
%Data are collected without focus on a group of diseases.
%
%This allows to query the data regarding many diseases and conditions.
%
%
\section{Prior and Related Work}
%% https://en.wikipedia.org/wiki/Exploratory_data_analysis#cite_note-Tukey1977-4
In 1977, Tukey already stated that data is too often only analyzed using confirmatory data analysis \cite{Tukey}.
%%
He emphasized the need to use data to derive hypotheses, which can then be tested again.
%%
In this section we present prior and related work trying to achieve this goal.
%%
We put each paper in relation to our work, derive ideas and point out differences, starting with the works closest to ours:
%%
\\\\
The work of Piringer et al. \cite{Piringer}, Chan et al. \cite{Chan}, Angelelli et al. \cite{Angelelli} and Zhang et al. \cite{Zhang2014} is closest to ours regarding different aspects.
%% HyperMoVal: Interactive Visual Validation of Regression Models for Real-Time Simulation
Akin to our approach, Piringer et al. \cite{Piringer} propose methods for visualizing regression analysis results and properties for developing car engines.
%%
Their main goal is to assess the influence of independent features towards the target pairwise.
%%
This is achieved by using a plot matrix displaying models as contours towards the target feature.
%%
They also incorporate 3D visualizations for each pairwise combination, but mainly because it is popular with the engineers in their target domains.
%%
Linked views of model deviations allow to select outlier residuals.
%%
Comparison of models is also included by mapping model values on line-style (e.g. solid for $model_a$ and dashed for $model_b$).
%%
This limits the method to comparing a few models at once.
%%
The plot matrix representation however gets complex and harder to understand with increasing number of independent features.
%%
The main difference is their focus on analyzing one complex model in detail, yielding in extensive plots, while we analyze a large amount of simpler models towards \emph{different} features.
%%
They also focus on metric features, while we process also categorical data.
%%
We share the matrix-arrangement of results depicting relationships between independent features in a model.

@@

Zhang et al. \cite{Zhang2014} is closest to our work regarding the application of visual analysis on cohort study data.
%%
They present \emph{Cohort Analysis via Visual Analytics} (\emph{CAVA}), a framework that distinguishes three major elements of cohort study data analysis: \emph{cohort data} and its manipulation using operations - \emph{views} and \emph{analytics}.
%%
They use the system to find longitudinal pathways for diseases using subject health records.
%%
Sub-groups are created automatically by applying \emph{analytics} and expert-guided \emph{views}.
%%
They do not go into the detail w.r.t the analytics used, they seem to not incorporate confounder analysis (other than dividing groups by gender) and the user study lacks concrete goals.
%%
Their data differs strongly from ours, having multiple moments for each subject, yielding a approach focused on finding pathways of interest.
%%
We use their requirements as guidelines, which incorporate \emph{flexible} and \emph{iterative analysis}, meaning that the system has to adapt to different requirements in different analysis steps.
%%
We implement this by supporting both hypothesis-based and hypothesis-free analysis.

Angelelli et al. \cite{Angelelli} visualize both image-derived an non image data using cube data structures with focus on comparison and knowledge extraction.
%%
Their proposed engine represents multiple heterogenous cohort study data sets as multiple normalized data cubes to eliminate data redundancy between them and perform runtime measure aggregation when measures of the different cubes have to be combined.
%%
%This allows for saving memory space.
%%
%Their presented aggregation functions allow for a fast data calculation to map it using different linked visualizations, such as scatter plots, bar charts or histograms.
%%
They use \emph{Pearsson's r} to depict relationships towards target features and employ list views and scatter plots to visualize and rank them.
%%
We use regression models and associated $R^2$ values to depict relationships between a target feature and multiple independent features instead of one and employ the information into one global visualization.
%%
We employ the idea of fast conducting large scale correlation analysis between many features to derive insight into the data.

\paragraph{Visual Analysis in Public Health.}
%%--
Shneiderman et al. \cite{Shneiderman2013} highlight the necessity of interactive visualizations in healthcare data, e.g. using \emph{Health 2.0}, which is an umbrella for web-based health-informatics frameworks.
%%
They divide Health 2.0 in three domains, \emph{Personal health information}, \emph{clinical health information}, and \emph{public health information}.
%%
We focus on the latter.
%%
%Shneiderman and colleagues also present challenges, 
The presented challenges in this application comprise custom tailoring the systems to be comprehensible and intuitive to fit into the short time span clinicians have when analyzing data.
%%
Furthermore, the systems should provide comparative relationship visualization, characterization of similarity and presenting risks.
%%
%We use the challenges as design guidelines for our web-based prototype, focusing on functional clean designs to allow fast insight into relationships towards diseases.

%%
In their survey on interactive information visualizations for health record data, Rind el al. \cite{Rind} identified future work addressing the usability of such frameworks.
%%
They recommend in opening up analysis systems to a diverse target group, e.g. including patients, who also can observe information about their own lifestyle and associated risks.
%%
We follow their recommendation in designing an open system available to everyone, applicable to a various data sets.

%%
Schreck and Keim \cite{Schreck} presented a method for analyzing social media data with focus on characterization and cause of an epidemic.
%%
They employ data aggregation techniques and use geographical data to map social media events, focusing on word clouds to derive insight into possible causes.
%%
They derive hypotheses using their high variation average message density algorithms, steering the users focus on particular events.
%%
%The focus remains on geolocated social media data rather than subjects with multiple dimensions.
%%
%These and similar systems follow the same analysis work as we do to derive insight into data.

%% Steenwijk: Integrated Visual Analysis for Heterogeneous Datasets in Cohort Studies
Steenwijk et al. \cite{Steenwijk} focus on \emph{hypothesis-free} exploration of cohort data.
%%
They employ a framework consisting of feature extraction and visualization to derive dependencies between image- and non-image features.
%%
They incorporate linked views using scatter plots, bar charts, parallel coordinates and time plots to display and select data.
%%
%Medical image data can be incorporated to investigate potentially interesting individuals.
%%
We distinguish to this and other linked-view analysis based approaches by showing abstracted information derived from a preceding analyze-first step to highlight relationships rather than displaying the actual data using various information visualization techniques.

>>>>

%\com{New Solution}.
%%
We propose an analysis approach of epidemiological data sets by incorporating all features in an exhaustive regression-based analysis.
%%
This approach combines all \emph{independent features} with respect to a \emph{target feature} and provides a visualization, which reveals insights into the data by highlighting relationships.
%%
A 3D-visualization of all combinations of two to three independent features towards a target acts as an overview of the whole data set, the \emph{Regression Cube}.
%%
Slicing through the \emph{Regression Cube} allows for the detailed analysis of features towards the target disease.
%%
Expert knowledge about disease-specific hypotheses can be included into the analysis by adjusting the regression model formulas.
%%
Furthermore, the influences of features can be assessed using a difference view comparing different calculation results.
%%
%\com{Validation \& Results}.
%%
We applied our \emph{Regression Cube} method to a hepatic steatosis data set to reproduce results from a data-mining driven analysis.
%%
A qualitative analysis with three domain experts was conducted on a breast fat data set.
%%
We were able to derive new hypotheses about relations between breast fat density and breast lesions towards breast cancer.
%%
%\com{Implications}.
%%
With our work, we present a visual overview of epidemiological data with the \emph{Regression Cube}, which allows for the first time an interactive regression-based analysis of large feature sets with respect to a disease.
} % end of abstract

@@

\maketitle
%%
Epidemiology aims to characterize health and disease conditions in defined populations (\emph{cohorts}).
%%
Insights about risk factors allow to characterize disease-specific high-risk groups and act as important diagnostic key figures \cite{Fletcher2012}.
%%
Furthermore, the insights can be used to derive recommendations regarding a healthy lifestyle or to provide information about wide-spread diseases.
%%
During the standard workflow, physicians transform observations into hypotheses, which are depicted using epidemiological features and then statistically analyzed.

@@

Testing features for association with diseases using regression analysis is one of the most important epidemiological tools.
%%
%Using it to assess the statistical resilience of a hypothesis rarely involves \emph{more than three features} due to the required subject count.
Using it to assess the statistical resilience of a hypothesis rarely involves \emph{more than three features} due to the higher dimensional problem and required subject count.
%%
% missing overview techniques ist zu hart; es gibt ja techniken (eine scatterplot matrix zb.)
Due to the amount of data and only limited overview techniques, possible correlations may be missed.
%%
Explorative analyses and overview visualizations of the data set as presented by Klemm et al. \cite{Klemm2014VIS} are not tailored to a specific target variable and mostly highlight correlations between variables, which are known to the domain expert (e.g., correlation between body size and spine shape).
%%
We incorporate the regression analysis, which is familiar to the domain experts, into overview visualizations to support a hypothesis-free analysis or an analysis towards a specific disease or hypothesis.
%%
This is achieved by providing template regression formulas, which are applied to all potential variable combinations.
%%
Since the notation is familiar to epidemiologists, they can rapidly include their domain knowledge into the analysis process.
%%
Difference views between regression formulas allow to assess the influences of individual variables on the process.

Our contributions are:
\begin{itemize}
	\item An overview visualization technique representing feature interactions using target features.
	\item Visualization techniques, which incorporate overview visualizations of all regression analyses at once as well as details on demand techniques for in-depth investigations of feature relationships.
	\item Freely adjustable regression formulas provide a simple, yet powerful way to adjust the regression analysis to specific hypotheses about the data.
	\item The analysis of confounding variables by providing comparison views between different formula results.
%	\item The open and web-based approach of the system allows for analysis of any data using the presented method.
	\item The expansion and adaption of our method to any data with our open and web-based system.
\end{itemize}
%%
The term Regression Cube has been used for the detailed analysis of one or a small number of regression models \cite{Chan, Ahmadi}.
%%
They extend 2D scatter plots to display detailed information about a regression model.
%%
In behalf of that, we declare our visualization technique for comparing and analyzing a large number of different regression models with different input features as Regression Cube.
%%
\section{Epidemiological Background} \label{sec:Background}
%%
This section covers the epidemiological workflow and requirements.
%%
\subsection{Epidemiological Workflow} \label{EpidemiologicalWorkflow}
Epidemiological research is performed by experts from different academic disciplines, such as physicians, statisticians and medical computer scientists focusing on biometrics and image segmentation.
%%
Their goal is to derive disease-specific risk factors by assessing epidemiological features using statistical methods.
%%
As described by Thew et al. \cite{Thew2009} the epidemiological workflow is divided into these different steps:
\begin{itemize}
	\item Clinicians make observations in the daily practice, which are translated into hypotheses.
	\item Epidemiologists compile a list of variables depicting the hypothesis and include confounding variables.
	\item Statisticians assess the association of the derived features towards the investigated disease.
\end{itemize}
%%
Relative risks can be determined if a statistical resilient association of features towards a condition is extracted.
%%
They indicate the per-subject chance of developing the disease.
%%
Reproducibility of results is an epidemiological key requirement and guides all analyses steps.
%%
Statistical programs, such as \texttt{SPSS}, are used to analyze the data using the classical sequential epidemiological workflow. Hence, images are mostly used for the communication of results, rather than for providing insight into the data.
%%

In \cite{Klemm2014VIS} we described an Interactive Visual Analysis approach for image-centric cohort study data, which connects to the variable listing step.
%%
The methods aim to derive hypotheses through analysis of the data and observations about previously unknown feature correlations.
%%
Employing \emph{hypotheses generation} requires overview visualizations of feature correlations, which are not supported by standard statistical processors.
%%
In this work, we focus on a similar approach to deriving insight and even new hypotheses through the data rather than using it just for confirmatory analysis.
%An alternative data-driven approach is described as follows.
%%
% \paragraph{Data Driven Hypothesis-Generation-based Analysis.}
% We already described an Interactive Visual Analysis approach for image-centric cohort study data, which connects to the variable listing step \cite{Klemm2014VIS}.
% %%
% The methods aim to derive hypotheses through analysis of the data and observations about previously unknown feature correlations.
% %%
% Employing \emph{hypotheses generation} requires overview visualizations of feature correlations, which are not supported by standard statistical processors.

\subsection{Epidemiological Data} \label{sec:EpidemiologicalData}
%%
%\com{Bernhard: Due to overlaps to VIS'14 paper in this section, this can be shortened. E.g. "More discussion of data types and characteristics can be found in [...]"}
%%
There are many data acquisition modalities available to epidemiologists, which are suitable for different analysis types.
%%
%We focus on cohort studies are a major epidemiological tool for gathering epidemiological features.
We focus on cohort studies, which impose the largest data sets and yield a highly heterogenous and incomplete information space.
%%
Subject data are collected either towards a specific disease or with the widest range possible.
%%
The latter allows the data set to be assessed towards different diseases.
%%
The feature space comprises information about lifestyle, somatometric variables, medical parameters, genetic data as well as medical image data.
%%
These features are derived through different modalities, such as questionnaires, medical examinations or laboratory analyses.
%%
Many features are sparse, such as follow up questions about a medication or treatment of a certain disease.
%%
Other features are exclusive for a sub-group, such as women-specific questions, e.g., number of born children or period status.
%%
%\com{Restriction of amount of data because it must not be triangulated and because the ethics committees are very restrictive}

%\paragraph{Data Types.}
%%
Medical status variables or lifestyle factors are often of \emph{dichotomous} (binary) type.
%%
The data space comprises continuous variables (somatometric variables, such as body weight) as well as categorical variables (e.g. graduation type).
%%
The data heterogeneity has to be taken into account when the analysis method is chosen.
%%
Continuous data are often discretized (e.g. 10 year steps for age) to equalize the variable types and to simplify the method selection.
%%
However, this should be avoided, since it reduces the information space and also introduces a new information bias, as assumptions are modeled through the discretization.
%%

Medical image data is also acquired and analyzed in modern cohort studies.
%%
We incorporate image-derived features, but do not put focus on analyzing image data.
%%
More discussion of cohort study data types and characteristics can be found in \cite{Klemm2014VIS, Toennies2015, Preim2014}.
%%
%\paragraph{Image Data.}
% \com{Can probably be cut (reference to Klaus' Paper for more details on this) (see Bernhards comment above)}
% %%
% Since the Rotterdam study, many modern cohort studies include medical image data.
% %%
% For ethical reasons, the imaging modalities usually not include ionizing radiation.
% %%
% The image quality is often inferior to clinical standards, which is a tradeoff between time and cost \cite{Preim2014}.
% %%
% These data are hard to analyze as they require segmentations to outline structures of interest.
% %%
% This process is prone to inter- and intra-observer variability when carried out manually.
% %%
% Automatic or semi-automatic solutions are a remedy for this problem, but are costly and need to be tailored to the structure of interest \cite{Toennies2015}.
% %%
% Image-derived variables are either dichotomous variables indicating a medical finding by a radiologist or continuous, describing the shape of segmented structures (e.g. volume or diameter).
\\\\
%\paragraph{Confounding Features.}
Features influencing the exposure as well as the outcome of an analysis are called \emph{confounders} and have to be specially treated.
%Confounding features have to be considered in epidemiological analysis, as they influence both the exposure as well as the outcome of an analysis.
%%
%Features influencing the exposure as well as the outcome of an analysis are called confounders.
%%
The analysis model has to be adjusted by normalizing all included features towards the confounder.
%%
\emph{Age} is included as confounder in almost any epidemiological analysis, since most diseases (such as different cancer types) are more likely with increasing age.
%%
It also influences the general body condition and thereby almost all features acquired through cohort studies.
%%
Another important confounder is \emph{Gender}.
%%
Other confounders have to be selected by epidemiologists specific to the investigated condition.

@@

\paragraph{Regression Analysis Notation.} Regression formulas are usually denoted as follows:
%%
\begin{equation}
Dependent \sim Independent_1 + ... + Independent_n
%Dependent \sim \sum_{i=1}^n Independent_i
\label{eq:RegressionNotation}
\end{equation}
%%
A example of a regression formula would be $Kidney\_Disorder ~ Smoking + Obesity$.
The most used regression operators comprise:
%%
\begin{itemize}
	\item $+,-$ inclusion/exclusion of the variable,
	\item $:$ inclusion of interactions between the variables (e.g. $x:y$),
	\item $*$ inclusion the variables as well as their interactions (e.g. $x*y$)
	\item $|$ (conditioning) inclusion of variable x, given y (e.g. $x|y$)
\end{itemize}
%%
The type of the target feature restricts the regression type.
%%
Different regression types are available, we focus on the following:
%%
\paragraph{Linear Regression for Continuous Target.} The basic type is the linear regression, creating a linear map from the space comprising the \emph{independent} features to the \emph{dependent} features.
%%
The \emph{dependent} variable has to be of a continuous type.
%%

\paragraph{Logistic Regression for Dichotomous Target.} Logistic regression implies a dichotomous target variable.
%%
The target is described by fitting a logistic function.
%%
Logistic models as opposed to linear models do not allow for extracting an $R^2$ quality of fit value.
%%
Therefore, pseudo-$R^2$ values are extracted, such as the \emph{Nagelkerke $R^2$}, which mimics the behavior of the $R^2$.
%%
\emph{Nagelkerke $R^2$} cannot be compared to $R^2$ values extracted from linear regression model.

@@

\subsection{The Study of Health in Pomerania (SHIP)}
The \texttt{SHIP}, located in Northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%%
It does not focus on a specific disease, making the data set open for many diseases.
%%
Unique for the \texttt{SHIP} is the acquisition of medical image data per cohort subject.
%%
A second cohort, \texttt{SHIP-TREND} was started in 2012.
%%
Data for both cohorts are examined in a 5-year time span.
%%
New parameters are added in each iteration, extending the range of investigated diseases.
%%
For the latest two acquisitions (\texttt{SHIP-2} and \texttt{SHIP-TREND-0}), MRI scans are included into the cohort \cite{Hegenscheid2009, Ivanovska2014}.
%%
However, most examinations occur in all stages and are performed according to the same instructions to enable comparisons over the different stages of the study ("\emph{method loyality}").
%%
%Starting in 1997 with a cohort of 4,308 subjects, the \texttt{SHIP}, located in Northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%After the pioneering Rotterdam study (started in 1990), several MR imaging study initiatives were initiated.
%
%They slightly differ in clinical focus, acquired data and epidemiological research questions.
%
%Starting in 1997 with a cohort of 4,308 subjects, the \texttt{SHIP}, located in Northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%
%Data are collected without focus on a group of diseases.
%
%This allows to query the data regarding many diseases and conditions.
%
%
\section{Prior and Related Work}
%% https://en.wikipedia.org/wiki/Exploratory_data_analysis#cite_note-Tukey1977-4
In 1977, Tukey already stated that data is too often only analyzed using confirmatory data analysis \cite{Tukey}.
%%
He emphasized the need to use data to derive hypotheses, which can then be tested again.
%%
In this section we present prior and related work trying to achieve this goal.
%%
%We put each paper in relation to our work, derive ideas and point out differences, starting with the works closest to ours:
%%
\\\\
The work of Piringer et al. \cite{Piringer}, Chan et al. \cite{Chan}, Angelelli et al. \cite{Angelelli} and Zhang et al. \cite{Zhang2014} is closest to ours regarding different aspects.
%% HyperMoVal: Interactive Visual Validation of Regression Models for Real-Time Simulation
Akin to our approach, Piringer et al. \cite{Piringer} propose methods for visualizing regression analysis results and properties for developing car engines.
%%
Their main goal is to assess the influence of independent features towards the target pairwise.
%%
This is achieved by using a plot matrix displaying models as contours towards the target feature.
%%
They also incorporate 3D visualizations for each pairwise combination, but mainly because it is popular with the engineers in their target domains.
%%
Linked views of model deviations allow to select outlier residuals.
%%
Comparison of models is also included by mapping model values on line-style. %(e.g. solid for $model_a$ and dashed for $model_b$).
%%
This limits the method to comparing a few models at once as the plot matrix gets complex with increasing feature number.
%%
%The plot matrix representation however gets complex and harder to understand with increasing number of independent features.
%%
The main difference is their focus on analyzing one complex model in detail, yielding in extensive plots, while we analyze a large amount of simpler models towards \emph{different} features.
%%
They also focus on metric features. %, while we process also categorical data.
%%
%We share the matrix-arrangement of results depicting relationships between independent features in a model.

@@

Zhang et al. \cite{Zhang2014} is closest to our work regarding the application of visual analysis on cohort study data.
%%
They present \emph{Cohort Analysis via Visual Analytics} (\emph{CAVA}), a framework that distinguishes three major elements of cohort study data analysis: \emph{cohort data} and its manipulation using operations - \emph{views} and \emph{analytics}.
%%
They use the system to find longitudinal pathways for diseases using subject health records.
%%
Sub-groups are created automatically by applying \emph{analytics} and expert-guided \emph{views}.
%%
%They do not go into the detail w.r.t the analytics used, they seem to not incorporate confounder analysis (other than dividing groups by gender) and the user study lacks concrete goals.
%%
%Their data differs strongly from ours, having multiple moments for each subject, yielding a approach focused on finding pathways of interest.
%%
We use their requirements as guidelines, which incorporate \emph{flexible} and \emph{iterative analysis}. %, meaning that the system has to adapt to different requirements in different analysis steps.
%%
%We implement this by supporting both hypothesis-based and hypothesis-free analysis.

Angelelli et al. \cite{Angelelli} visualize both image-derived an non image data using cube data structures with focus on comparison and knowledge extraction.
%%
Their proposed engine represents multiple heterogenous cohort study data sets as multiple normalized data cubes to eliminate data redundancy between them and perform runtime measure aggregation when measures of the different cubes have to be combined.
%%
%This allows for saving memory space.
%%
%Their presented aggregation functions allow for a fast data calculation to map it using different linked visualizations, such as scatter plots, bar charts or histograms.
%%
They use \emph{Pearsson's} $r$ to depict relationships towards target features and employ list views and scatter plots to visualize and rank them.
%%
%We use regression models and associated $R^2$ values to depict relationships between a target feature and multiple independent features instead of one and employ the information into one global visualization.
%%
We employ the idea of fast conducting large scale correlation analysis between many features to derive insight into the data.

\paragraph{Visual Analysis in Public Health.}
%%--
Shneiderman et al. \cite{Shneiderman2013} highlight the necessity of interactive visualizations in healthcare data. %, e.g. using \emph{Health 2.0}, which is an umbrella for web-based health-informatics frameworks.
%%
%They divide Health 2.0 in three domains, \emph{Personal health information}, \emph{clinical health information}, and \emph{public health information}.
%%
%We focus on the latter.
%%
%Shneiderman and colleagues also present challenges, 
The presented challenges for \emph{public health information visualization} comprising custom tailoring the systems to be comprehensible and intuitive to fit into the short time span clinicians have when analyzing data.
%%
Furthermore, the systems should provide comparative relationship visualization, characterization of similarity and presenting risks.
%%
%We use the challenges as design guidelines for our web-based prototype, focusing on functional clean designs to allow fast insight into relationships towards diseases.
%%
%%
In their survey on interactive information visualizations for health record data, Rind el al. \cite{Rind} identified future work addressing the usability of such frameworks.
%%
They recommend in opening up analysis systems to a diverse target group, e.g. including patients, who also can observe information about their own lifestyle and associated risks.
%%
%We follow their recommendation in designing an open system available to everyone, applicable to a various data sets.
%%
%%
Schreck and Keim \cite{Schreck} presented a method for analyzing social media data with focus on characterization and cause of an epidemic.
%%
They employ data aggregation techniques and use geographical data to map social media events, focusing on word clouds to derive insight into possible causes.
%%
They derive hypotheses using their high variation average message density algorithms, steering the users focus on particular events.
%%
%The focus remains on geolocated social media data rather than subjects with multiple dimensions.
%%
%These and similar systems follow the same analysis work as we do to derive insight into data.
